{\plant}: A package for modelling forest trait ecology \& evolution

### 2015-05-29 ms.tex
Further tweaks to main text

Although the importance of size-structure for vegetation dynamics and
diversity, has been long been recognised \citep[e.g.,][]{Harper-1977,
  Shugart-1980, Huston-1987}, current research in this area is
dominated by models and theory that either omit size-structure
entirely, or only include coarse approximations. 
%
Theoretical investigations exploring questions about niche-based
species coexistence and ecological drift (neutrality) both primarily
rely on unstructured population models \citep[e.g.][]{MacArthur-1967,
  Tilman-1985, Geritz-1995, Hubbell-2001, Calcagno-2006}.
%
It is common to assume competitive interactions are influenced by
size-related traits such as adult size and seed size, but detailed
size-structure is rarely considered, at least for plant models
\cite[for animal examples, see][]{Deroos-1988, Deroos-1992}.
% TODO: I don't have the background to write it, but (at the risk of
% making a long paragraph even longer) is it worth adding something
% saying something about how we'd expect size structure to be
% particularly important for plants because they grow up through a
% canopy?  Perhaps not useful?
Consequently, it has been difficult to connect and test the mechanisms
described in these simplistic models with types of data collected in
the field, such as physiological leaf traits or demographic data from
long-term monitoring plots \citep{Adler-2013}.
%
Similarly, most large-scale models of vegetation dynamics do not
include size-structured demography \citep{Cramer-2001, Sitch-2003,
  Kelley-2013,Dekauwe-2014}.  During the 1980s, size-structured
``gap'' models were widely used
\citep[e.g.][]{Shugart-1980,Huston-1987}. However, the next generation
of models focused more heavily on biogeochemistry at the expense of
demography \citep[e.g.][]{Woodward-1995,Haxeltine-1996}. While the
importance of size-structure is starting again to be recognised
\citep{Moorcroft-2001, Falster-2011, Dekauwe-2014}, all but a couple
\citep[e.g.][]{Moorcroft-2001, Smith-2014} of the most popular,
large-scale vegetation models lack any real demography or
size-structure.

In this note, we introduce the {\plant} package for R \citep{R-2015};
a mechanistic framework for studying the effects of size structure and
trait variation on the demography of individual plants, of patches of
competing plants and of meta-populations structured by a prevailing
disturbance regime.
% TODO: the first sentence reads weirdly to me: the paragraph now says
% "we introduce our model / other models are like X / our model is
% also like X" I'm not sure what the key point that is being driven at
% here, either.
%
% TODO: I've reordered it and changed the emphasis here.  Can you
% check that you agree and if so add a reference.
We developed {\plant} primarily to investigate if and how species
differing in traits are able to coexist with one another
\citep{Falster-2011,Falster-2015}.  In order to allow interactions to
be driven by species traits and arise through competition for shared
resources, we needed to model a community with a full range of sizes,
as species may not be able to recruit under the canopy that they
generate (REF).

At the same time, we expect the {\plant} model to be useful for
modelling other phenomena. The internal structure of the {\plant}
model is similar in approach to the ``Ecosystem Demography''
\citep{Moorcroft-2001} and \textsc{lpj-guess} \citep{Smith-2014}
models. Like {\plant}, these models include size- structured
competition within a metapopulation of ``patches''. However, whereas
these models were designed to capture large-scale regional
biogeochemistry driven by climatic variables and rely on simulations,
{\plant} was designed so that the rules driving the dynamics of the
system could be manipulated and emergent demographic behaviours
studied in detail.  In {\plant} it is possible to grow plants in
isolation, in patches of competing plants and in disturbance-driven
metapopulations. The package is thus designed to sit in the middle
ground between a large-scale coupled simulation model and simple
abstract theoretical device. 
% TODO: Consider moving to the conclusion?
As examples, we expect the package will prove useful for investigating
how morphological and physiological traits influence growth rate and
shade tolerance, and emergent phenomena such as multi-species
self-thinning, successional transitions and age-related changes in
vegetation productivity.

% TODO: they aren't actually available within plant yet, and
% realistically won't be by the updated.  I've added a placeholder
% that contains appropriate links.
Below we describe the general approach of the package, then provide
sections with details at nested levels of ecological organisation.
Detailed technical documents are provided as supplementary information
(see Appendices), and updated versions of technical documents are also
available within the {\plant} package itself.

@@

Second, the metapopulation concept allows for the effects of disturbance
regimes on vegetation structure and dynamics to be properly incorporated. In
{\plant} characteristics for the entire vegetation (metapopulation) -- such
as total leaf area, biomass and productivity -- are obtained by averaging
patch-level outcomes over the frequency-distribution of patch ages. Because
young patches are more common in the landscape, metapopulation averages are
strongly influenced by the structure of younger patches (e.g. 
Fig. \ref{fig:patch}c). Moreover, predictions are for a frequency distribution 
of states across the landscape. Similarly, the distributions of  abundance and growth rate
observed in large forest  plots \citep[e.g.][]{Muller-2006, Coomes-2007} may
can be modelled as  emergent phenomena, where variety of patch-level
relationships combine to  produce  an emergent average across all patches
(Fig. \ref{fig:emergent}, see more detail in Appendix \ref{sec:emergent}). As such,
size-and patch- structured models may be more suitable for comparing with 
satellite and survey data, than unstructured models representing the world as 
homogeneous green slime \citep{Moorcroft-2001,Moorcroft-2001}.

Finally, with the extension to metapopulation, we are modelling demography across
the entire life cycle and are thus in a position to estimate the fitness
of different types. Invasion fitness quantifies the ability of a rare
individual with particular traits to establish in a community of
established residents at their equilibrium densities \citep{Metz-1992}.
Invasion fitness
is ideally calculated as the long-term per capita growth rate of the new
type; however, in structured metapopulations, a more convenient
indicator of per-capita growth rate is given by the logarithm of the
basic reproduction ratio \citep{Gyllenberg-2001, Metz-2001}. The basic
reproduction ratio \(R\) is simply the total seed output of the new
type, averaged across the metapopulation. A new type can invade when
\(R > 1\).

With fitness defined, it becomes possible to model adaptive dynamics
and community assembly \citep{Geritz-1998,Dieckmann-2007}. This is best
thought of as occurring on a fitness landscape, a surface with fitness
plotted against one or more species traits (Fig. \ref{fig:fitness}). The
reproductive success of individuals possessing a particular trait value
is measured in the size- and patch-structured metapopulation and
depends on both the traits and densities of other species in the current
resident community. The reason this is interesting is that under
frequency- and density-dependence, evolution tends to favour different
trait mixtures compared to those predicted by models that optimise some
feature of the community, such as total biomass or carbon production or
seed production in the absence of competition
\citep{Falster-2003, Dieckmann-2007}.

Fitness landscapes will be positive in regions that allow invasion,
and zero for species that are at equilibrium. The slope of the fitness
landscape at a resident indicates the direction of selection; a
negative slope (dashed line in Fig. \ref{fig:fitness}a) indicates
selection for smaller trait values (see Appendix \ref{sec:fitness} for
details). Fitness landscapes are constantly reshaped as species are
introduced and evolve, reflecting the non-linear demographic feedbacks
that occur under frequency-dependent evolution
\citep{Geritz-1998,Dieckmann-2007}. Following the fitness gradient,
can lead to an evolutionary branching point
\citep{Geritz-1998,Dieckmann-2007}; at this point there is no
directional selection (solid line in Fig. \ref{fig:fitness}a) but
invasion is possible both above and below it in trait space. It is
often possible to find a configuration where fitness is negative
everywhere (solid line in Fig. \ref{fig:fitness}b) except for the zero
points where species are present. This community is immune to invasion
from any other strategies \citep{Geritz-1998}.

Identifying conditions promoting trait diversity is a valuable use-case
for {\plant}. Coexisting plant species differ in a range of physiological
traits, yet the conditions allowing for these different types to coexist
in face of competition for a common set of resources remain largely
unknown. The simple example here can be extended to multiple dimensions
and can be used to investigate how trade-offs facilitate coexistence
\citep{Falster-2015}

@@

The {\plant} package makes it easier for researchers and forest practitioners
alike to investigate a variety of non-linear demographic phenomena.  Rather
than provide a model with a few entry points, we have developed an extensible
framework that we hope can be used to answer a number of questions at a
variety of scales. The specific use-cases highlighted here arise from our
interest in the assembly of trait mixtures in communities. However a variety
of other uses would be possible and also extensions to address other aspects
of community assembly. For example,  the physiology of growth could be
modified to account for rising atmospheric CO2 or the effects of competition
for multiple resources could likewise be investigated.

It is hoped that the approach here can be used to inform both future empirical
research and model development in neighbouring areas. As noted above, most
large-scale mechanistic models of plant growth do not explicitly model size
structure, so remain unable to investigate demographic phenomena
\citep{Cramer-2001, Sitch-2003, Kelley-2013,Dekauwe-2014}. At the same time,
much attention is paid to detailed leaf-level physiology. {\plant} aims to bridge
this gap that emerged in the model landscape, where detailed physiology and
size-structured dynamics combine to drive forest structure and function via
traits.

@@

DS Falster was supported by an ARC discovery grant (DP110102086). RG
FitzJohn was supported by the Science and Industry Endowment Fund
(RP04-174). M Westoby was supported by a fellowship from Australian
Research Council. We thank {\AA} Brn{\"a}nnstr{\"o}m, J Camac, 
A De Roos, U Dieckmann, R Duursma, and J Johansson for helpful discussions.

>>>>

Although the importance of size-structure for vegetation dynamics and
diversity, has been long been recognised \citep[e.g.,][]{Harper-1977,
  Shugart-1980, Huston-1987}, current research in this area is
dominated by models and theory that either omit size-structure
entirely, or only include coarse approximations. 
%
Theoretical investigations exploring questions about niche-based
species coexistence and ecological drift (neutrality) both primarily
rely on unstructured population models \citep[e.g.][]{MacArthur-1967,
  Tilman-1985, Geritz-1995, Hubbell-2001, Calcagno-2006}.
%
It is common to assume competitive interactions are influenced by
size-related traits such as adult size and seed size, but detailed
size-structure is rarely considered, at least for plant models
\cite[for animal examples, see][]{Deroos-1988, Deroos-1992}.
%
Consequently, it has been difficult to connect and test the mechanisms
described in these simplistic models with types of data collected in
the field, such as physiological leaf traits or demographic data from
long-term monitoring plots \citep{Adler-2013}.
%
Similarly, most large-scale models of vegetation dynamics do not
include size-structured demography \citep{Cramer-2001, Sitch-2003,
  Purves-2008,Dekauwe-2014}.  During the 1980s, size-structured
``gap'' models were widely used
\citep[e.g.][]{Shugart-1980,Huston-1987}. However, the next generation
of models focused more heavily on biogeochemistry at the expense of
demography \citep[e.g.][]{Woodward-1995,Haxeltine-1996,Cramer-2001, 
Sitch-2003}. While the
importance of size-structure is starting again to be recognised
\citep{Moorcroft-2001, Purves-2008, Falster-2011, Dekauwe-2014}, all but a couple
of the most popular,
large-scale vegetation models lack any real demography or
size-structure \citep[e.g.][]{Moorcroft-2001, Smith-2014}.

In this note, we introduce the {\plant} package for R \citep{R-2015};
a mechanistic framework for studying the effects of size structure and
trait variation on the demography of individual plants, of patches of
competing plants and of meta-populations structured by a prevailing
disturbance regime.
% 
We developed {\plant} primarily to investigate if and how species
differing in traits are able to coexist with one another
\citep{Falster-2011,Falster-2015}.  In order to allow interactions to
be driven by species traits and arise through competition for shared
resources, we needed to model a community with a full range of sizes,
as species may not be able to recruit under the canopy that they
generate \citep{Moorcroft-2001,Huston-1987}.

At the same time, we expect the {\plant} model to be useful for
modelling other phenomena. The internal structure of the {\plant}
model is similar in approach to the ``Ecosystem Demography''
\citep{Moorcroft-2001} and \textsc{lpj-guess} \citep{Smith-2014}
models. Like {\plant}, these models include size- structured
competition within a metapopulation of ``patches''. However, whereas
these models were designed to capture large-scale regional
biogeochemistry driven by climatic variables and rely on simulations,
{\plant} was designed so that the rules driving the dynamics of the
system could be manipulated and emergent demographic behaviours
studied deterministically and in detail. In {\plant} it is possible to grow plants in
isolation, in patches of competing plants and in disturbance-driven
metapopulations. Moreover, we put considerable effort into resolving
competitive hierarchies. 
The package is thus designed to sit in the middle
ground between a large-scale coupled simulation model and simple
abstract theoretical device.

Below we describe the general approach of the package, then provide
sections with details at nested levels of ecological organisation.
Detailed technical documents are provided as supplementary information
(see Appendices), updated versions of these technical documents will also
be made available within the {\plant} package itself.

@@

Second, the metapopulation concept allows for the effects of disturbance
regimes on vegetation structure and dynamics to be properly incorporated. In
{\plant} characteristics for the entire vegetation (metapopulation) -- such
as total leaf area, biomass and productivity -- are obtained by averaging
patch-level outcomes over the frequency-distribution of patch ages. Because
young patches are more common in the landscape, metapopulation averages are
strongly influenced by the structure of younger patches (e.g. 
Fig. \ref{fig:patch}c).  Similarly, the distributions of  abundance and growth rate
observed in large forest  plots \citep[e.g.][]{Muller-2006, Coomes-2007} arise
as  emergent phenomena. Moreover, the predictions from {\plant} 
are for a frequency distribution of patch states across the landscape 
(Fig. \ref{fig:emergent}, see more detail in Appendix \ref{sec:emergent}). As such,
size-and patch- structured models may be more suitable for linking with 
satellite and forest survey data, than unstructured models representing the 
world as homogeneous green slime \citep{Moorcroft-2001, Purves-2008}.

Finally, with the extension to metapopulation, we are modelling demography across
the entire life cycle and are thus in a position to estimate invasion fitness -- 
the ability of a rare
individual with particular traits to establish in a community of
established residents at their equilibrium densities \citep{Metz-1992}.
Invasion fitness
is ideally calculated as the long-term per capita growth rate of the new
type; however, in structured metapopulations, a more convenient
indicator of per-capita growth rate is given by the logarithm of the
basic reproduction ratio \citep{Gyllenberg-2001, Metz-2001}. The basic
reproduction ratio \(R\) is simply the total seed output of the new
type, averaged across the metapopulation. A new type can invade when
\(R > 1\).

With fitness defined, it becomes possible to model adaptive dynamics
and community assembly \citep{Geritz-1998,Dieckmann-2007}. This is best
thought of as occurring on a fitness landscape, a surface with fitness
plotted against one or more species traits (Fig. \ref{fig:fitness}). The
reproductive success of individuals possessing a particular trait value
depends on both the traits and densities of other species in the current
resident community. Fitness landscapes will be positive in regions that allow invasion,
and zero for species that are at equilibrium. The slope of the fitness
landscape at a resident indicates the direction of selection; a
negative slope (dashed line in Fig. \ref{fig:fitness}a) indicates
selection for smaller trait values (see Appendix \ref{sec:fitness} for
details). Fitness landscapes are constantly reshaped as species are
introduced and evolve, reflecting the non-linear demographic feedbacks
that occur under frequency-dependent evolution
\citep{Geritz-1998,Dieckmann-2007}. Following the fitness gradient,
can lead to an evolutionary branching point
\citep{Geritz-1998,Dieckmann-2007}; at this point there is no
directional selection (solid line in Fig. \ref{fig:fitness}a) but
invasion is possible both above and below it in trait space. It is
often possible to find a configuration where fitness is negative
everywhere (solid line in Fig. \ref{fig:fitness}b) except for the zero
points where species are present. This community is immune to invasion
from any other strategies and represents a prediction for the 
trait mixture that might might be favoured by natural selection \citep{Geritz-1998}.

Identifying conditions promoting trait diversity is a valuable use-case
for {\plant}. Coexisting plant species differ in a range of physiological
traits, yet the conditions allowing for these different types to coexist
in face of competition for a common set of resources remain largely
unknown. The simple example here can be extended to multiple dimensions
and can be used to investigate how trade-offs facilitate coexistence
\citep{Falster-2015}.

@@

The {\plant} package makes it easier for researchers and forest practitioners
alike to investigate a variety of non-linear demographic phenomena.  Rather
than provide a model with a few entry points, we have developed an extensible
framework that we hope can be used to answer a number of questions at a
variety of scales. The specific use-cases highlighted here arise from our
interest in the assembly of trait mixtures in communities. However a variety
of other uses would be possible and also extensions to address other aspects
of community assembly. As examples, we expect the package will prove useful
for investigating how morphological and physiological traits influence growth
rate and shade tolerance, and emergent phenomena such as multi-species self-
thinning, successional transitions and age-related changes in vegetation
productivity. The effects of competition
for multiple resources could likewise be investigated.

It is hoped that the approach here can be used to inform both future empirical
research and model development in neighbouring areas. As noted above, most
large-scale mechanistic models of plant growth do not explicitly model size
structure, so remain unable to investigate demographic phenomena
\citep{Cramer-2001, Sitch-2003, Dekauwe-2014}. At the same time,
much attention is paid to detailed leaf-level physiology. {\plant} aims to bridge
this gap that emerged in the model landscape, where detailed physiology and
size-structured dynamics combine to drive forest structure and function via
traits.

@@

DS Falster was supported by an ARC discovery grant (DP110102086). RG
FitzJohn was supported by the Science and Industry Endowment Fund
(RP04-174). M Westoby was supported by a fellowship from Australian
Research Council. We thank {\AA} Brn{\"a}nnstr{\"o}m, J Camac, 
A De Roos, U Dieckmann, R Duursma, J Johansson and C Prentice 
for helpful discussions.
